---
title: "Gov 50 Final Project"
author: "Heidi Enger"
description: "The Effect of Post-Shelby County vs. Holder Voting Laws on Voter Turnout"
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

For this project, I was interested in exploring data related to U.S. elections. I chose to examine voter turnout in states which passed restrictive voting laws in the aftermath of the 2013 Shelby County v. Holder Supreme Court case, which found Section 4 of the Voting Rights Act unconstitutional, allowing states to pass voting laws without being subject to federal review. Restrictive laws include restrictions on mail-in voting and unreasonably strict voter ID requirements, which disproportionately impact voters of color. These sorts of laws have been adopted by 29 states as of 2023 (Brennan Center for Justice, 2023).

Did restrictive state voting laws after Shelby County vs. Holder (2013) impact voter turnout in the 2016 general election? In this report, I examine the extent to which restrictive voting laws passed after the Shelby decision impacted voter turnout in applicable states. I hypothesize that voter turnout decreased from 2012 to 2016 in states which passed stricter voter ID and mail-in ballot laws, especially for voters of color. I've chosen a data set with election data from 1948-2020, though I'll only be looking at general election data from 2012 and 2016, since I want to compare across time. The explanatory variable of interest is whether a state passed restrictive voting law(s) or not. I had to filter to specific states from the data based on information from the Brennan Center (see above). My outcome variable is voter turnout. If I observe lower voter turnout in 2016 compared to 2012 in states which passed restrictive voting laws, this would provide support for my hypothesis. If I observe higher or similar voter turnout, this would provide evidence against my hypothesis. 

## The Data 
The data source is the American National Election Studies (ANES) Time Series Cumulative Data File, 1948-2020. ANES has conducted surveys, generally in-person interviews, during most national election cycles since 1948. The key dependent and independent variables for my project, namely voter turnout and year, are measured by these surveys. My research design is cross-sectional (difference-in-differences?)

First, I filtered the dataset to only 2012 and 2016, directly before and after the Shelby County vs. Holder decision. Then I selected the columns I thought would be useful by referencing the code book, and renamed them for clarity. 

```{r}
library(tidyverse)
elections <- read_csv("anes_timeseries.csv", show_col_types = FALSE)|>
  filter(VCF0004 == 2012 | VCF0004 == 2016) |>
  select(VCF0004, VCF0101, VCF0104, VCF0105a, VCF0112, VCF0113, VCF0301, VCF0310, VCF0701, VCF0702, VCF0703, VCF0748, VCF0750, VCF0803, VCF0901a, VCF9150) |>
  rename(year = VCF0004, age = VCF0101, gender = VCF0104, race = VCF0105a, region = VCF0112, poli_south = VCF0113, party = VCF0301, interest = VCF0310, registered = VCF0701, voted = VCF0702, turnout_summary = VCF0703, vote_day = VCF0748, vote_method = VCF0750, lib_scale = VCF0803, state_fips = VCF0901a)

```
To summarize the dependent variable, voter turnout, I created a plot of overall voter turnout in the 2012 vs. 2016 national elections:

```{r}
elections <- elections |>
  filter(voted == 2 | voted == 1) |>
  mutate(voted_or_not = if_else(voted == 2, 1, 0))

turnout_props <- elections |>
  group_by(year) |>
  summarize(turnout_prop = mean(voted_or_not))

knitr::kable(turnout_props, col.names = c("Year", "Turnout Proportion"))

overall_turnout <- elections |>
  group_by(year) |>
  summarize(turnout_prop = mean(voted_or_not)) |>
  ggplot(mapping = aes(y = turnout_prop)) +
  geom_col(aes(x = year), fill = "blue", width = 2) +
  ylim(0,1) +
  labs(x = "Year", y = "Turnout Proportion", title = "Overall voter turnout in 2012 vs. 2016") +
  annotate(geom = "text", x = 2012, y = 0.83, label = "Prop = 0.799") +
  annotate(geom = "text", x = 2016, y = 0.81, label = "Prop = 0.786")

overall_turnout
```
As you can see from the plot, overall voter turnout decreased from 2012 to 2016. But I wanted to examine the effects in states previously subject to preclearance that enacted restrictive voting laws after Shelby County vs. Holder. Those states are South Dakota, New York, North Carolina, South Carolina, Georgia, Florida, Alabama, Mississippi, Louisiana, Texas, and Arizona. The FIPS codes for those states are 1, 4, 12, 13, 22, 28, 36, 37, 45, 46, and 48. 

# Results

```{r}
library(ggplot2)

elections <- elections |>
  mutate(new_laws = if_else(state_fips == c(1, 4, 12, 13, 22, 28, 36, 37, 45, 46, 48), "Restrictive laws", "No new laws"))

turnout_props_laws <- elections |>
  group_by(new_laws) |>
  summarize(turnout_prop = mean(voted_or_not))

knitr::kable(turnout_props_laws, col.names = c("New laws or not", "Turnout Proportion"))

laws_comparison <- elections |>
  filter(year == 2016) |>
  group_by(new_laws) |>
  summarize(turnout_prop = mean(voted_or_not)) |>
  ggplot(mapping = aes(y = turnout_prop)) +
  geom_col(aes(x = new_laws), fill = "indianred") +
  labs(x = "Whether state passed new restrictive laws", y = "Turnout proportion", title = "Turnout in states that passed restrictive laws vs. states with no new laws") 

laws_comparison

```
The turnout proportion in states that passed new restrictive voting laws was lower than the turnout in states that did not. *More interpretation here*

plot of main analysis + regression output + 2-3 paragraphs of description and interpretation of the plots and regression (including interpreting the main coefficient of interest and describing if it is statistically significant and if we should interpret it causally). 

```{r}

```

You should interpret (in plain English) the main coefficient of interest in your regression. You should also comment on the statistical significance of the estimated coefficient and whether or not you believe the coefficient to represent a causal effect.

# Conclusion
a brief (one paragraph) concluding section that summarizes your results, 
assesses the extent to which you find support for your hypothesis, 
describes limitations of your analysis and threats to inference, 
states how your analysis could be improved (e.g., improved data that would be useful to collect).

Possible confounding: fewer people voting in 2016 due to unusually divisive candidates: Hilary Clinton, a woman, and Donald Trump, a political outsider. 

## Works Cited
American National Election Studies. 2022. ANES Time Series Cumulative Data File 1948-2020 [dataset and documentation]. September 16, 2022 version. https://electionstudies.org/data-center/anes-time-series-cumulative-data-file/

Singh, Jasleen, and Sara Carter. “States Have Added Nearly 100 Restrictive Laws Since SCOTUS Gutted the Voting Rights Act 10 Years Ago.” Brennan Center for Justice, 23 June 2023, https://www.brennancenter.org/our-work/analysis-opinion/states-have-added-nearly-100-restrictive-laws-scotus-gutted-voting-rights.



